A Core Calculus of Metaclasses

### 2004-12-16 metaclasses/paper.tex
submitted version of paper

\documentclass{acm-sigplan}

@@

\begin{abstract}
Metaclasses provide a useful mechanism for abstraction in
object-oriented languages.  But most languages supporting metaclasses
have imposed severe restrictions on their use. Typically, a metaclass
is allowed to have only a single instance and all metaclasses are
required to share a common superclass \cite{objvlisp}. In addition,
few languages supporting metaclasses have included a static type
system, and none have included a type system with nominal subtyping
(i.e., subtyping as defined in mainstream languages such as the
Java${^{\mbox{\tiny{TM}}}}$ Programming Language or C\#).

In order to elucidate the structure of
metaclasses and their relationship with static types, we present a
core calculus for a nominally-typed object-oriented language with
metaclasses and prove type soundness over this core. 
To our knowledge, no previous formalization of metaclasses in a 
language with nominal subtyping exists.
This calculus is
presented as an adaptation of Featherweight GJ \cite{FJ}, and is
powerful enough to capture metaclass relationships beyond what are
expressible in common object-oriented languages, including arbitrary
metaclass hierarchies and classes as values.  We also explore the
integration of object-oriented design patterns with metaclasses, and
show how our calculus integrates several such patterns cleanly.
\end{abstract}

\section{Introduction}
One of the stated benefits of object-oriented languages is their
ability to model aspects of the world in the object hierarchy.
However, most such languages are unable to model many simple and
common relationships.

An excellent example is provided by Welty and Ferrucci in \cite{FW96}, where
they show the difficulties in modeling a simple
ontology including the notions {\txt{Species}} and {\txt{Eagle}}, while
also capturing the relationship between {\txt{Eagle}} and {\txt{Harry}},
a particular eagle.

To see the problem, consider a system with a class \txt{Eagle} and an
instance \txt{Harry}.  The static type system can ensure that any
messages that \txt{Harry} responds to are also understood by any other
instance of \txt{Eagle}.

However, even in a language where classes themselves can have methods,
such relationships cannot be expressed.  For example, in a language
with static methods, we can write the expression
\txt{Eagle.isEndangered()}

However, if \txt{Salmon} is a species, our type system provides us no
way of requiring that it also have an \txt{isEndangered} method.  The
solution that Ferruci and Welty suggest, and that we develop in this
paper, is to allow a consistent treatment of both instances such as
\txt{Harry} and classes such as \txt{Eagle}.

One solution they consider and reject is to separate the notion of
``eagle'' into two concepts: {\txt{the Eagle}} and {\txt{Eagle}}, with
the former an instance of {\txt{Species}} and the latter a class whose
instances are specific birds.  This approach is inadequate because it
fails to express a relationship between the notions {\txt{the Eagle}}
and {\txt{Eagle}}.  Therefore, if we create another species such as
\txt{Salmon} in our universe, we must define one new class and one new
instance.  This adds new invariants which the programmer must
maintain, with no help from the type system.  Instead, we propose that
the type system should allow the programmer to express the desired
relationship directly, and enforce the invariants naturally.

For example, consider the relationship between
physical quantities, units of measurement, and physical dimensions. It
is obvious that ``length'' is a dimension, and that ``3 feet'' is a
length.  However, modeling this relationship in a conventional
language such as the Java Programming Language is difficult without
attributing to some of these concepts properties which do not properly
belong to them.  It is natural to model {\txt{3 feet}} as an instance
of a class {\txt{Length}}. But if we were to define a class {\txt
Dimension} and define {\txt{Length}} to be an instance of class {\txt
Dimension}, then {\txt{Length}} could not be a class and so {\txt 3
feet} could not be an instance of {\txt{Length}}. Alternatively, we
might define {\txt{Length}} to be a subclass of {\txt{Dimension}}. But
then we still cannot define {\txt 3 feet} to be an instance of {\txt
Length} because anything that is an instance of {\txt{Length}} would
also be an instance of class {\txt{Dimension}} and {\txt 3 feet} is
obviously not a dimension. The key limitation preventing us from
modeling this domain properly is that we cannot define a single
concept to serve both as a class and as an instance of a class.

Again, the limitation is that we cannot define a class as an instance of another class.
%Welty and Ferrucci proceed to explain how this limitation prevents conventional modeling
%languages from capturing many important aspects of the world.  
%The languages Ferrucci and Welty discuss cannot accommodate multiple
%universes of discourse, in which a single thing may have different meanings in different universes.

In order to overcome this limitation in expressiveness, several
object-oriented languages that are not statically typed, such as
Smalltalk, Python, and Self, allow for more flexible object
relationships \cite{Smalltalk80}, \cite{ArtMOP}, \cite{PythonMeta}. In
the case of Smalltalk and Python, classes are instances of
metaclasses. Static members of a class are modeled as ordinary members
of the corresponding metaclass.  But in both of these languages there
are important limitations on expressiveness.  For example, in
Smalltalk, the metaclass hierarchy is only two levels deep.  If we
want to represent the relationships ``{\txt A} is an instance of {\txt
B}", ``{\txt B} is an instance of {\txt C}", and ``{\txt C} is an
instance of {\txt D}", we can only do this in Smalltalk if {\txt D} is
the special Smalltalk class {\txt Metaclass}.  This does not allow the
multi-level hierarchies required to model the examples discussed above
\cite{Smalltalk80}.  Self, on the other hand, is a prototype-based
object-oriented language, where there are no classes at all; object
instantiation consists of cloning an existing instance. The members of
the clone can be modified and added to at will, dispensing with class
relationships entirely and making static checking difficult.

A more expressive metaclass system in ObjVlisp, described informally
by Cointe in \cite{objvlisp}.  In ObjVlisp, there is no restriction on
the number of instances of a metaclass, and metaclasses are not
required to share a common superclass. The instance methods of a
metaclass are inherited as the static members of its instances.  Conte
makes a number of arguments for the benefits of generalized
metaclasses.  In response to complaints that the Smalltalk80 metaclass
system is too complex, he suggests that a more powerful and flexible
system has the potential to be much simpler.
%, arguing that ``The ObjVlisp model's primary advantage is uniformity.''
However, ObjVlisp does not include a static type system. 
%Cointe quotes from \cite{cardelli:multiple}: ``With respect to
%Simula, Smalltalk also abandons ... strong typing, allowing it to
%... introduce the notion of meta-classes''.   
In this paper, we show that this limitation is not necessary. We
present a semantics for a statically typed calculus of metaclasses
where, as in ObjVlisp, every class can serve as both an instance of a
class and as a class of instances, and there is no limitation on the
levels of nesting of metaclasses. In the process, we also reveal
several interesting properties about the structure that a nominal type
system for metaclasses must take. We then present a formal operational
semantics for this calculus, as well as a proof of type soundness.

Our work is motivated by \cite{DimUnits}, which presents a system for
integrating the static checking of dimensions of physical quantities
in MetaGen, an extension of the Java Programming Language with support
for metaclasses. In MetaGen, metaclasses are used to model the type
relationships of dimensions.  As with ObjVlisp, the notion of a
metaclass in MetaGen is more general than that of languages such as
Smalltalk, but, unlike ObjVlisp, MetaGen includes a static type system
\cite{DimUnits}. The calculus in this paper can be viewed as a
formalization of a purely functional core of MetaGen.

@@

%Another, somewhat more flexible solution to this problem can be found in
%functional languages, where sets can be represented as predicates over
%objects, and sets of sets can be represented as predicates over
%predicates.  This hierarchy can obviously be extended indefinitely,
%but such programs would quickly become unwieldy.  Specialized language
%constructs would be needed for this approach to become a serious programming
%style, just as most object-oriented programming languages
%provide constructs to support that methodology.  Also in the
%functional world, higher-order functions represent a way of
%parameterizing at multiple levels, allowing some of the desired
%behavior described here to be implemented.  But few functional
%programs nest higher-orderness more than a few levels, since
%understanding the relationships  becomes more than one programmer
%could follow.  Instead, we need a way of structuring such
%relationships explicitly.

%Formal modeling of existing languages has become common, with numerous
%formal models for languages such as the Java (TM) Programming Language.  However, formal modeling
%can also serve to aid in the design process for a new language.  In
%this paper, we describe MCJ, a nominally-typed core calculus for metaclasses,
%which is intended to be the basis for a language design. While we have focused on a formal model as a language design guide, MCJ can also help understand existing languages with metaclasses.

@@

MCJ is a object-oriented calculus with generic types based on
Featherweight GJ which includes metaclasses. An MCJ program consists of
a sequence of class definitions followed by a trailing
expression. This trailing expression is evaluated in the context of
the class table induced by the class definitions.  Before discussing
the formal specification of MCJ, we first give a high-level overview
of the nature of MCJ class definitions.

@@

\begin{verbatim}
class C kind D extends Object {...}
\end{verbatim}
\footnote{In other examples, we will leave out the \kind\ and \extends\
declaration when they refer to \Object.}

Each class may have both class members and instance members. Class
members define the behavior of the class when used as an
instance. Instance members define the behavior of instances of a
class.  In the abstract syntax presented in this paper, class members
are distinguished from instance members by order, and with the keyword
\class \footnote{This syntax is not ambiguous because MCJ has no
inner classes.}.  Members of a class definition are laid out as
follows: first, class fields are defined, followed by class methods,
then instance fields and finally instance methods.  For example, the
following class definition includes a class method {\txt m} and an
instance method {\txt n}:

\begin{verbatim}
class C {
   class Object m() {...}
   Object n() {...}
}
\end{verbatim}

Class fields and methods bear a resemblance to static fields and
methods, but there are important differences. For a given class {\txt
C}, instance members of {\txt C}'s kind are inherited as class
members of {\txt C}. For example, a class method {\txt m} in {\txt C}
with the same name as an instance method of {\txt C}'s kind
overrides the definition of {\txt m} in the kind (and it must have the
same signature as the overridden method). Note that if an instance
class {\txt C} of type {\txt T} is assigned to a variable {\txt v} of
type {\txt T}, references to the instance members of {\txt v} refer to
the class members of {\txt C}.  As in \cite{DimUnits}, a class is
allowed to define both a class method and an instance method of the
same name. We prevent ambiguity in member references by requiring that
all references to class members in MCJ must explicitly denote the
receiver.

Superclasses behave as in \FGJ, providing implementation inheritance
of instance behavior and subtyping. A class, when used as an
expression, is an instance of its kind.  If a method is invoked on a
class {\txt C}, first the class methods of {\txt C} are examined.  If
the method is found there, it is invoked. Otherwise, the
\emph{instance} methods of the kind {\txt D} of {\txt C} are searched,
and if it is found there, the method is invoked.  Otherwise, the
superclass hierarchy of {\txt D} is examined as it is for an ordinary
instance of {\txt D}.

@@

To resolve the method call $\invoke \C {\mof {}}$ first the class
methods of $\C$ are examined, and if a match is found, it is invoked.
Then the \emph{instance} methods of $\B$ and then $\A$ are examined.

@@

%Class {\txt C} would first be examined to see if it contained a class method named {\txt m}. If so, this method would be invoked. Otherwise, class {\txt D} would be checked to see if it contained an instance method named {\txt m}. If not, class {\txt E} would then be checked for an instance method named {\txt m}.

The generic type system of MCJ is similar to the generic type system
in \FGJ. However, as in \cite{NextGen, DimUnits}, we allow for type
variables to occur in type-dependent contexts such as casts,
preventing the use of type erasure as an implementation technique. In
addition, the receiver of a reference to a class member may be a type
variable. Unlike the system presented in \cite{MixGen}, MCJ does not
support first-class genericity because a naked type variable must not
appear in the \extends\ clause or \kind\ clause of a class definition
and {\txt{new}} expressions on naked type variables are not
supported.\footnote{First-class genericity is difficult to integrate
with our presentation of constructors.} Also, polymorphic methods
(which are orthogonal to the features we explore) are not supported.

Because classes can be used as expressions, we need a bound on the
behavior of classes when used in expression contexts.  Therefore, we
place two bounds on every type variable {\txt T} in the header of the
class definition. The first bound on {\txt T} is a bound on the kind
of an instantiation of {\txt T}; the second bound is a bound on the
superclass of {\txt T}. For example, the following class header
declares a class {\txt C} with one type parameter {\txt T} that must
be instantiated with a type of kind {\txt D} that is a subtype of
{\txt E}:

@@

One benefit of a metaclass system is that constructors no longer need
to play such a central role in the language. Class instances exist at
the beginning of a program execution; they need not be
constructed. Non-class instances are naturally constructed with class
(factory) methods \cite{GOF}; in this way a class instance can be
passed as a type parameter and factory methods can be called on the
class instance to construct new instances without stipulating the
class of the constructed instance. This flexibility is similar to that
provided with first-class genericity \cite{NextGen, MixGen}, but
obviates the need for {\txt{with}} clauses because the bound on the
kind of a type variable stipulates what factory methods can be called
on it.

For example, in MixGen the following program is legal:

@@

Here, no with clause is required.

A class can define class methods with arbitrary signatures that return
new instances of the class, but each such method must ultimately
include a {\txt{new}} expression, which can occur only within the
scope of the class whose instance it returns. A {\txt{new}} expression
looks like a call to method named {\txt{new}}, that takes a single
argument for each type parameter and each instance field of the
syntactically enclosing class. The result of a {\txt{new}} expression
is a new instance of the enclosing class whose type parameters are
instantiated with, and whose fields are initialized with, the given
arguments. In our formal semantics, {\txt{new}} expressions are
annotated with the name of the enclosing class. These annotations need
not be added by a programmer; they can be added easily by a
straightforward syntactic preprocessing over the program, since they
can be determined merely by the lexical scopes of the {\txt{new}}
expressions.

All uses of constructors in \FGJ\ \cite{FJ} are macro-expressible
\cite{FellExpress} in MCJ, that is, they can be expressed via local
transformations.  Of course, not all \FGJ\ programs can be expressed
in MCJ, because of the differences in the type system. Namely, MCJ
does not include polymorphic methods.

Metaclasses allow other kinds of flexibility in object creation, as
discussed in section \ref{designpatterns}.

@@

However, classes which are themselves instances of other classes have
fields as well, and these fields must also be initialized before they
are used.  These fields include those instance fields of the kind
which become class fields of the new class.  Our solution is to
require that \emph{all class fields, including those obtained by
inheritance, must be redeclared} and provided with values.

% In a {\txt{new}} expression, all instance fields must be initialized at
% creation time. This requirement eliminates the problem of
% uninitialized instance fields. However, when a class is the kind of
% another class, the instance fields of the kind become class fields of
% the instance class.  Since the instance class can then be used as a
% value with no further initialization, these variables must be
% initialized somewhere. Our solution is to require that every class
% field inherited from a container class be included in a new class
% field definition and that every class field in a class definition
% include an initializing expression.

@@

Given that class references can be used as expressions, it is natural
to ask: what is the type of a class reference?  
%In the world of the
%$\lambda$-calculus, such types are known as kinds.  
In MCJ, however, the kind of a class does not capture all of the
properties of the class as a value.  For example, a class may add new
class fields or class methods which are not present in the kind.
Therefore, each class freely generates a new type, which is the type
of the class considered as a value.  We represent this type with the
{\txt{typeOf}} operator, which takes a class instantiation and produces
a type that is not also a class.  Because this type is not a class, it
cannot be the superclass or kind of another class, and it cannot serve
as the instantiation of a generic type parameter.  We refer to this as
{\txt{typeOf}} because it is the type of the expression that is its
argument.

@@

Another difference between subclassing and {\txt{instanceof}}
relationships is that cycles can occur among instance class
relationships. The simplest such cycle is:

@@

Although this pathology appears to be dangerously close to Russell's
paradox, it does not lead to inconsistencies.  Class \C\ is an
instance of itself. It can be used in any context where one of its
instances can be used.  \C\ can be thought of as a self-replicating
value. This class hierarchy causes no problems for method lookup
because method resolution never proceeds through more than a single
containing class to an instance class. If the programmer uses a field
of \C\ to initialize a class field of \C\, the evaulation of the
expression \C\ may fail to terminate.  But this is no different than
the possibility of non-termination from any other self-reference.
Therefore, we see no reason to disallow it.

@@

Having seen this outline of MCJ, we return to the motivational
examples discussed earlier.  First, the relationships between
dimensions is easy to capture. \footnote{Of course, dimensions have
many subtleties, not captured here.  All we present is the essence of
the type relationships.} We define

@@

Here {\txt{Meter}} is a singleton class (there is only one Meter, in
the Platonic sense).  We have easily expressed the desired type
relationships, and can statically check program invariants that rely
on dimensional relationships.

@@

\begin{verbatim}
class Species { ... }
class Eagle kind Species {
      class Eagle make(String name) { ... }
}
Eagle harry = Eagle.make("Harry")
\end{verbatim}

@@

Another problem, which is in some ways just a symptom of the first
problem but which bears special attention, is the use of generic
types.  Consider the following class definition, added to the
definitions of \txt{Species}, \txt{theEagle}, etc. above.

\begin{verbatim}
class C<T extends X> {
   foo(T x) { return T.isEndangered() }
}
\end{verbatim}

If we want to perform the call {\txt{foo(harry)}}, then we have two
choices for the bound \txt{X} on {\txt T}.  It must be either \Object,
in which case {\txt{m}} must be a method that \Object\
understands. Otherwise, it must be \txt{Eagle}, in which case the
function cannot handle multiple {\txt{Species}}.  The ``solution'' in
conventional languages is to use a bound of \Object\ and insert a
downcast, which fails at runtime if the wrong argument is passed.
This is the problem that generics were intended to alleviate.  In
MCJ {\txt T} can be bound by kind {\txt{Species}}, and instantiated
with {\txt{Eagle}}, allowing the original call to be type-checked.
This gives the programmer both safety and expressiveness.

\paragraph{Metaclasses and Nested Classes}
Some of the problems are alleviated in a language with nested, or
inner classes.  In the language Scala\cite{Scala} the following
program expresses some of the desired invariants:

\begin{verbatim}
class Species {
  class Member {}
}
object Eagles extends Species;
class Eagle() extends Eagles.Member;
val harry = new Eagle;
\end{verbatim}
\footnote{This example, and this discussion, was suggested by an
anonymous reviewer.}

This program automates many of the invariants required in a language
without metaclasses.  However, two seperate classes must still be
constructed per \txt{Species}.  Further, the fundamental problem, of
not being able to group classes by functionality in the same way as
instances, still remains.  Inner classes are not constrained by
invariants imposed by their container.  In MCJ, the invariants of the
kind are enforced by the type system.

@@

Design Patterns, as exemplified in \cite{GOF}, have had a substantial
impact on the world of object-oriented programming and beyond.  Design
patterns allow programmers to increase the flexibility and abstraction
of their software designs.  However, few design patterns have been
integrated into programming languages.

>>>>

\documentclass[nocopyrightspace,10pt]{acm-sigplan}

@@

\begin{abstract}
Metaclasses provide a useful mechanism for abstraction in
object-oriented languages.  But most languages that support metaclasses
 impose severe restrictions on their use. Typically, a metaclass
is allowed to have only a single instance and all metaclasses are
required to share a common superclass \cite{objvlisp}. In addition,
few languages that support metaclasses include a static type
system, and none include a type system with nominal subtyping
(i.e., subtyping as defined in languages such as the
Java${^{\mbox{\tiny{TM}}}}$ Programming Language or C\#).

To elucidate the structure of metaclasses and their relationship with
static types, we present a core calculus for a nominally typed
object-oriented language with metaclasses and prove type soundness
over this core.  To our knowledge, no previous formalization of
metaclasses in a language with nominal subtyping exists.  This
calculus is presented as an adaptation of Featherweight GJ \cite{FJ},
and is powerful enough to capture metaclass relationships beyond those
expressible in common object-oriented languages, including arbitrary
metaclass hierarchies and classes as values.  We also describe how the
addition of metaclasses allow for integrated and natural expression of
several common design patterns.
\end{abstract}

\section{Introduction}

\subsection{A Problem of Modeling}
One of the stated benefits of object-oriented languages is their
ability to model aspects of the world in the object hierarchy.
However, most such languages are unable to model many simple and
common relationships.  An excellent example is provided by Welty and
Ferrucci \cite{FW96}, where they show the difficulties in modeling
a simple ontology including the notions {\txt{Species}} and
{\txt{Eagle}}, while also capturing the relationship between
{\txt{Eagle}} and {\txt{Harry}}, a particular eagle.  To see the
problem, consider a system with a class \txt{Eagle} and an instance
\txt{Harry}.  The static type system can ensure that any messages that
\txt{Harry} responds to are also understood by any other instance of
\txt{Eagle}.

However, even in a language where classes can have static methods, the
desired relationships cannot be expressed.  For example, in such a
language, we can write the expression \linebreak[4]\txt{Eagle.isEndangered()}
provided that \txt{Eagle} has the appropriate static method.  Such a
method is appropriate for any class that is a species.  However, if
\txt{Salmon} is another class in our system, our type system provides
us no way of requiring that it also have an \txt{isEndangered}
method.  This is because we cannot classify both \txt{Salmon} and
\txt{Eagle}, \emph{when treated as receivers of methods}, into a
larger group, such as species.

One solution they consider and reject is to separate the notion of
``eagle'' into two concepts: {\txt{the Eagle}} and {\txt{Eagle}}, with
the former an instance of {\txt{Species}} and the latter a class whose
instances are specific birds.  This approach is inadequate because it
fails to express a relationship between the notions {\txt{the Eagle}}
and {\txt{Eagle}}.  Therefore, if we create another species such as
\txt{Salmon} in our universe, we must define one new class and one new
instance.  This adds new invariants which the programmer must
maintain, with no help from the type system.

These problems are not unique to the modeling of biology.  Consider
the relationship between physical quantities, units of measurement,
and physical dimensions. It is obvious that ``length'' is a dimension,
and that ``3 feet'' is a length.  However, modeling this relationship
in a conventional language such as the Java Programming Language is
difficult without attributing to some of these concepts properties
which do not properly belong to them.  It is natural to model {\txt{3
    feet}} as an instance of a class {\txt{Length}}. But if we were to
define a class {\txt Dimension} and define {\txt{Length}} to be an
instance of class {\txt Dimension}, then {\txt{Length}} could not be a
class and so {\txt 3 feet} could not be an instance of {\txt{Length}}.
Alternatively, we might define {\txt{Length}} to be a subclass of
{\txt{Dimension}}. But then we still cannot define {\txt 3 feet} to be
an instance of {\txt Length} because anything that is an instance of
{\txt{Length}} would also be an instance of class {\txt{Dimension}}
and {\txt 3 feet} is obviously not a dimension.

The fundamental problem that both these examples illustrate is that we
are unable to describe multi-level \txt{instanceof} hierarchies.
Everything must be either a class or an instance.  Classes cannot be
instances of classes, and instances cannot have instances.  The
solution to this problem is to remove these divisions, and allow
the relationships we want to have a natural expression in our
language.

\subsection{Some Previous Solutions}
In order to overcome this limitation in expressiveness described
above, several object-oriented languages that are not statically
typed, such as Smalltalk, Python, and Self, allow for more flexible
object relationships \cite{Smalltalk80}, \cite{ArtMOP},
\cite{PythonMeta}. In the case of Smalltalk and Python, classes are
instances of metaclasses. Static members of a class are modeled as
ordinary members of the corresponding metaclass.  But in both of these
languages there are important limitations on expressiveness.  For
example, in Smalltalk, the metaclass hierarchy is only two levels
deep.  If we want to represent the relationships ``{\txt A} is an
instance of {\txt B}'', ``{\txt B} is an instance of {\txt C}'', and
``{\txt C} is an instance of {\txt D}'', we can only do this in
Smalltalk if {\txt D} is the special Smalltalk class {\txt Metaclass}.
This does not allow the multi-level hierarchies required to model the
examples discussed above \cite{Smalltalk80}.  Self, on the other hand,
is a prototype-based object-oriented language, where there are no
classes at all; object instantiation consists of cloning an existing
instance.  The members of the clone can be modified and added to at
will, dispensing with class relationships entirely and making static
checking difficult.

A more expressive metaclass system is ObjVLisp, described informally
by Cointe \cite{objvlisp}.  In ObjVLisp, there is no restriction on
the number of instances of a metaclass, and metaclasses are not
required to share a common superclass. The instance methods of a
metaclass are inherited as the static members of its instances.  Cointe
makes a number of arguments for the benefits of generalized
metaclasses.  In response to complaints that the Smalltalk-80 metaclass
system is too complex, he suggests that a more powerful and flexible
system has the potential to be much simpler.
However, ObjVLisp does not include a static type system.  As Cardelli
points out in the context of Smalltalk \cite{cardelli:multiple}, ``With
respect to Simula, Smalltalk also abandons ... strong typing, allowing
it to ... introduce the notion of meta-classes''.

\subsection{A More Expressive Solution}

In this paper, we show that the limitations of previous metaclass
systems are not inherent. We present a semantics for a calculus with
metaclasses in which, as in ObjVLisp, every class can serve as both an
instance of a class and as a class of instances, and there is no
limitation on the levels of nesting of metaclasses. But, unlike
ObjVLisp, we present a static type system for this semantics, and use
it to allow more invariants to be checked statically.  In the process,
we also reveal several interesting properties about the structure that
a nominal type system for metaclasses must take. We then present a
formal operational semantics for this calculus, as well as a proof of
type soundness.

Our work is motivated by \cite{DimUnits}, which presents a system for
integrating the static checking of dimensions of physical quantities
in MetaGen, an extension of the Java Programming Language with support
for metaclasses. In MetaGen, metaclasses are used to model the type
relationships of dimensions.  As with ObjVLisp, the notion of a
metaclass in MetaGen is more general than that of languages such as
Smalltalk, but, unlike ObjVLisp, MetaGen includes a static type system
\cite{DimUnits}. The calculus in this paper can be viewed as a
formalization of a purely functional core of MetaGen.

@@

@@

MCJ is a object-oriented calculus with generic types based on
Featherweight GJ that includes metaclasses. An MCJ program consists of
a sequence of class definitions followed by a trailing
expression. This trailing expression is evaluated in the context of
the class table induced by the class definitions.  Before discussing
the formal specification of MCJ, we first give a high-level overview
of the nature of MCJ class definitions.

@@

\begin{verbatim}
class C kind D extends Object {...}
\end{verbatim}

Each class may have both static members and instance members. Static
members define the behavior of the class when used as an instance.
Instance members define the behavior of instances of a class.  In the
abstract syntax presented in this paper, static members are
distinguished from instance members by order, and with the keyword
\static.  Members of a class definition are laid out as follows:
first, static fields are defined, followed by static methods, then
instance fields and finally instance methods.  For example, the
following class definition includes a static method {\txt m} and an
instance method {\txt n}.  Beginning with this example, we will leave
out the \kind\ and \extends\ declaration when they refer to \Object.

\begin{verbatim}
class C {
   static Object m() {...}
   Object n() {...}
}
\end{verbatim}

Static fields and methods in MCJ bear a strong resemblance to static
fields and methods, but there are important differences. For a given
class {\txt C}, instance members of {\txt C}'s kind are inherited as
class members of {\txt C}. For example, a static method {\txt m} in
{\txt C} with the same name as an instance method of {\txt C}'s kind
overrides the definition of {\txt m} in the kind (and it must have the
same signature as the overridden method). Note that if an instance
class {\txt C} of type {\txt T} is assigned to a variable {\txt v} of
type {\txt T}, references to the instance members of {\txt v} refer to
the static members of {\txt C}.  As in \cite{DimUnits}, a class is
allowed to define both a static method and an instance method of the
same name. We prevent ambiguity in member references by requiring that
all references to static members in MCJ must explicitly denote the
receiver.

Superclasses behave as in \FGJ, providing implementation inheritance
of instance behavior and subtyping. A class, when used as an
expression, is an instance of its kind.  If a method is invoked on a
class {\txt C}, first the static methods of {\txt C} are examined.  If
the method is found there, it is invoked. Otherwise, the
\emph{instance} methods of the kind {\txt D} of {\txt C} are searched,
and if it is found there, the method is invoked.  Otherwise, the
superclass hierarchy of {\txt D} is examined as it is for an ordinary
instance of {\txt D}.

@@

To resolve the method call $\invoke \C {\mof {}}$, first the class
methods of $\C$ are examined, and if a match is found, it is invoked.
Then the \emph{instance} methods of $\B$ and then $\A$ are examined.

@@

%Class {\txt C} would first be examined to see if it contained a static method named {\txt m}. If so, this method would be invoked. Otherwise, class {\txt D} would be checked to see if it contained an instance method named {\txt m}. If not, class {\txt E} would then be checked for an instance method named {\txt m}.

The generic type system of MCJ is similar to the generic type system
in \FGJ. However, as in \cite{NextGen, DimUnits}, we allow for type
variables to occur in type-dependent contexts such as casts,
preventing the use of type erasure as an implementation technique. In
addition, the receiver of a reference to a static member may be a type
variable. Unlike the system presented in \cite{MixGen}, MCJ does not
support first-class genericity because a naked type variable must not
appear in the \extends\ clause or \kind\ clause of a class definition
and {\txt{new}} expressions on naked type variables are not
supported.\footnote{First-class genericity is difficult to integrate
with our presentation of constructors.} Also, polymorphic methods
(which are orthogonal to the features we explore) are not supported.

Because classes can be used as expressions, we need a bound on the
behavior of classes when used in expression contexts.  Therefore, we
place two bounds on every type variable {\txt T} in the header of the
class definition. The first bound on {\txt T} is a bound on the kind
of an instantiation of {\txt T}; the second bound is a bound on the
superclass of {\txt T}. For example, the following class header
declares a class {\txt C} with one type parameter {\txt T} that must
be instantiated with a type that is a subtype of {\txt D} and is of
kind {\txt E} :

@@

One benefit of a metaclass system is that constructors no longer need
to play such a central role in the language. Class instances exist at
the beginning of a program execution; they need not be constructed.
Non-class instances are naturally constructed with class (factory)
methods \cite{GOF}; in this way a class instance can be passed as a
type parameter and factory methods can be called on the class instance
to construct new instances without stipulating the class of the
constructed instance.  In a language with first-class genericity, type
parameters may used in arbitrary contexts, including in \txt{new}
expressions \cite{NextGen, MixGen}, however additional bounds, called
\txt{with} clauses, are required on the type parameters.  MCJ provides
similar flexibility, but obviates the need for {\txt{with}} clauses
because the bound on the kind of a type variable stipulates what
factory methods can be called on it.

For example, in MixGen \cite{MixGen} the following program is legal:

@@

Here, no \txt{with} clause is required.

A class can define static methods with arbitrary signatures that return
new instances of the class, but each such method must ultimately
include a {\txt{new}} expression, which can occur only within the
scope of the class whose instance it returns. A {\txt{new}} expression
looks like a call to method named {\txt{new}}, that takes a single
argument for each type parameter and each instance field of the
syntactically enclosing class. The result of a {\txt{new}} expression
is a new instance of the enclosing class whose type parameters are
instantiated with, and whose fields are initialized with, the given
arguments. In our formal semantics, {\txt{new}} expressions are
annotated with the name of the enclosing class. These annotations need
not be added by a programmer; they can be added easily by a
straightforward syntactic preprocessing over the program, since they
can be determined merely by the lexical scopes of the {\txt{new}}
expressions.

All uses of constructors in \FGJ\ \cite{FJ} are macro-expressible
\cite{FellExpress} in MCJ; that is, they can be expressed via local
transformations.  Of course, not all \FGJ\ programs can be expressed
in MCJ, because of the differences in the type system. Specifically,
MCJ does not include polymorphic methods.

While these choices may seem constricting, the architecture of MCJ
allows for considerable flexibility in object creation.  We discuss
this point further in section \ref{designpatterns}.

@@

Classes that are themselves instances of other classes have fields as
well, and these fields must also be initialized before they are used.
These fields include instance fields of the kind, which become static
fields of the new class.  Our solution is to require that \emph{all
  static fields, including those obtained by inheritance, must be
  redeclared} and provided with values.

% In a {\txt{new}} expression, all instance fields must be initialized at
% creation time. This requirement eliminates the problem of
% uninitialized instance fields. However, when a class is the kind of
% another class, the instance fields of the kind become static fields of
% the instance class.  Since the instance class can then be used as a
% value with no further initialization, these variables must be
% initialized somewhere. Our solution is to require that every class
% field inherited from a container class be included in a new class
% field definition and that every static field in a class definition
% include an initializing expression.

@@

Given that class references can be used as expressions, it is natural
to ask: what is the type of a class reference?  In MCJ, however, the
kind of a class does not capture all of the properties of the class as
a value.  For example, a class may add new static fields or static
methods that are not present in the kind.  Therefore, each class
freely generates a new type, which is the type of the class considered
as a value.  We represent these types with a compile time
{\txt{typeOf}} operator that takes a class instantiation and produces
a new type that is not also a class.  Because this type is not a class, it
cannot be the superclass or kind of another class, and it cannot serve
as the instantiation of a generic type parameter.  We refer to it as
{\txt{typeOf}} because it returns the type of the expression that is its
argument.

@@

Another difference between subclassing and instance class
relationships is that cycles can occur among instance class
relationships. The simplest such cycle is:

@@

Although this pathology appears to be dangerously close to Russell's
paradox, it does not lead to inconsistencies.  Class \C\ is an
instance of itself. It can be used in any context where one of its
instances can be used.  \C\ can be thought of as a self-replicating
value. This class hierarchy causes no problems for method lookup
because method resolution never proceeds through more than a single
containing class to an instance class. If the programmer uses a field
of \C\ to initialize a static field of \C\, the evaluation of the
expression \C\ may fail to terminate.  But this is no different than
the possibility of nontermination from any other self-reference.
Therefore, we see no reason to disallow it.

@@

Having seen this outline of MCJ, we return to the motivational
examples discussed earlier.  First, the relationships between
dimensions is easy to capture.\footnote{Of course, dimensions have
many subtleties, not captured here.  All we present is the essence of
the type relationships.} We define

@@

Here {\txt{Meter}} is a singleton class (there is only one Meter, in
the platonic sense).  We have easily expressed the desired type
relationships, and can statically check program invariants that rely
on dimensional relationships.

@@

\begin{verbatim}
class Species { ... }
class Eagle kind Species {
      static Eagle make(String name) { ... }
}
Eagle harry = Eagle.make("Harry")
\end{verbatim}

@@

Another problem, which is in some ways just a symptom of the first
problem but that bears special attention, is the use of generic
types.  Consider the following class definition, added to the
definitions of \txt{Species}, \txt{theEagle}, etc. above, where
\txt{?} is a placeholder for the bound.

\begin{verbatim}
class C<T extends ?> {
   T foo(T x) { return T.isEndangered() }
}
\end{verbatim}

If we want to perform the call {\txt{c.foo(harry)}}, then we have two
choices for the bound \txt{X} on {\txt T}.  It must be either \Object,
in which case {\txt{isEndangered}} must be a method that \Object\
understands. Otherwise, it must be \txt{Eagle}, in which case the
function cannot handle multiple {\txt{Species}}.  The ``solution'' in
conventional languages is to use a bound of \Object\ and insert a
downcast, which fails at runtime if the wrong argument is passed.
This is the problem that generics were intended to alleviate.  In
MCJ, {\txt T} can be bound by kind {\txt{Species}}, and instantiated
with {\txt{Eagle}}, allowing the original call to be type-checked.
This gives the programmer both safety and expressiveness.

@@

Design patterns, as exemplified in \cite{GOF}, have had a substantial
impact on the world of object-oriented programming and beyond.  Design
patterns allow programmers to increase the flexibility and abstraction
of their software designs.  However, few design patterns have been
integrated into programming languages.

One of the advantages of our metaclass framework is that it allows
clean expression of several common design patterns in the language,
rather than requiring that they be expressed with abstraction
techniques on top of the language.  For example, a number of the
classic design patterns deal with object creation, as discussed above
in section \ref{privconst}.  Several of these are naturally expressed
in MCJ.

\paragraph{Factory}
The requirement that all constructors be private enforces the factory
pattern for all MCJ programs.
\begin{verbatim}
class MyClass {
    static MyClass make() { return new(); }
}
\end{verbatim}
Since \txt{make()} is a method, any computation can be performed,
without the problems of constructors.

\paragraph{Abstract Factory}
 The Abstract Factory pattern is a simple application of
inheritance in the metaclass hierarchy.  In this, several classes
could share a kind, giving them all a single interface for
construction. Again, the example is very simple:
\begin{verbatim}
class AbsFac {
    AbsFac make() { ... }
}
class Derived kind AbsFac {
    AbsFac make() { return new(); }
}
\end{verbatim}
